Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment

### 2021-09-28 manuscript.Rmd
Manuscript Revision (minor addition)

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did predict different levels of contamination with highest probability among floors followed be elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Additionally, it is possible that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

@@

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. This finding matches previous description of high probability of floor contamination relative to other surfaces[@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

>>>>

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did predict different levels of contamination with highest probability among floors followed be elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Alternatively, it is possible that the quality of cleaning declined as the study progressed, or that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

@@

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. This finding matches previous description of high probability of floor contamination relative to other surfaces[@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets or aeresols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.
